Estimating Uncertainty for Tipping Points with Deep Learning

### 2022-05-11 manuscript/manuscript.Rmd

Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis].
Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will be necessary to use probablisitic forecasts. 
For all the deep learning models in this study, we use the same method for uncertainty estimation:
each model is trained to learn the parameters of a Poisson distribution to maximize the log likelihood. 
To compute uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.

@@

Recurrent neural networks are the predominant memory-based ML method.
Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps.
By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus are considered a memory-based method.
However, a standard recurrent neural network, i.e. a feed forward neural network that provides self-feedback, is unwieldy to train because of the vanishing and exploding gradient problem [@Pascanu], so there have been specialized neural network architectures designed to avoid these gradient problems.
Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung].
These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung for more details.
While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung], so we investigate both methods.

@@

The Transformer is a state of the art ML architecture which uses an attention mechanism to model long and short term dependencies on sequence to sequence tasks [@vaswani].
Specifically, Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani]. 
Because of self-attention, Transformers do not need to process data sequentially like recurrent neural networks do, and thus can be parallelized, offering significant computational advantages [@vaswani].
The Transformer is likely to be a foundational method for future AI research [@bommasani], so we considered it critical to investigate these models here.

@@

Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].

>>>>

Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@makridakis].
Deep learning models are often used to make point forecasts, yet for their application to ecological time series, it will often be necessary to use multi-step, probablisitic forecasts. 
For all the deep learning models in this study, we use the same method for uncertainty estimation:
each model is trained to learn the parameters of a Poisson distribution to maximize the log likelihood. 
To compute uncertainty intervals, we draw samples from the learned distribution at each time step and report the 95% confidence interval.

@@

Recurrent neural networks are the predominant memory-based ML method.
Recurrent neural networks differ from feed-forward neural networks in that a recurrent neural network provides feedback to itself between time steps [@sherstinsky].
By providing self-feedback, recurrent neural networks are able to retain information from previous time steps and thus are considered a memory-based method.
However, a standard recurrent neural network is unwieldy to train because of the vanishing and exploding gradient problem [@Pascanu], so there have been specialized neural network architectures designed to avoid these gradient problems.
Long Short-term Memory and Gated Recurrent Units Networks are considered to be the state of the art recurrent neural networks that address exploding and vanishing gradients [@chung].
These methods avoid gradient problems by regulating the self-feedback between time steps via gates which perform operations on the feedback signal -- see @chung for more details.
While GRU's and LSTM's commonly outperform standard RNN's, it is difficult to anticipate whether GRU's or LSTM's will be best suited for any time series problem [@chung], so we investigate both methods.

@@

The Transformer is a state of the art ML architecture that is able to model long and short term dependencies on sequence to sequence tasks [@vaswani].
Transformers use a mechanism called self-attention which interrelates different positions of the input sequence in order to find an informative representation of the input sequence [@vaswani].
For example, if given a sentence, a transformer could learn the contextual relationship between a subject and a direct object, but a recurrent network would process all the words as one phrase.
Because of self-attention, Transformers do not need to process data sequentially, and thus can be parallelized, offering significant computational advantages [@vaswani].
The Transformer is likely to be a foundational method for future AI research [@bommasani], so we considered it critical to investigate Transformers in this study.

@@

Encoder-decoder architectures have been shown empirically to excel on sequence to sequence tasks [@learningphrase, @cite, @cite].
Encoder-decoders work by processing the input sequence into a fixed-length vector then decoding the fixed-length vector to the predicted output sequence.
It is thought that by encoding the input sequence to a vector, encoder-decoders find informative representations of the input sequence that make the prediction task easier for the decoder [@sutskever].
Of the models that we present, Block RNNs and Transformers are the only models that use encoder-decoders.
